Modeling tabular icebergs coupled to an ocean model

### 2017-02-14 Icebergs_only.tex
More changes to iceberg only version

\begin{abstract}
The calving of giant icebergs from the Antarctic ice shelves accounts for approximately half of Antarctic ice-shelf decay. After calving these tabular icebergs drift across large distances, altering regional ocean circulation, bottom water formation and local rates biological primary productivity. Freshwater fluxes from sub-marine melting can affect large-scale sea-ice distributions, which feedback onto the greater climate system.
However, despite their importance, the current generation of numerical models is unable to represent giant tabular icebergs a physically realistic way.
In this study we develop novel method to model giant tabular icebergs. Tabular icebergs are constructed out of Lagrangian elements which are bonded together by numerical bonds. 
These numerical bonds hold the ice element together and allow the individual elements to move as a unit. By breaking these bonds, we can simulate the  ocean response to giant iceberg detaching away from an ice shelf (also constructed from Lagrangian elements) or the fracturing of a large iceberg into smaller pieces.  This paper provides a description and demonstration of the Lagrangian tabular-iceberg model and discusses a number of the technical elements involved its development. 
%We demonstrate the capabilities of the model by modeling a tabular iceberg breaking away an idealized ice shelf.
\end{abstract}
%% ------------------------------------------------------------------------ %%
%
%  BEGIN ARTICLE
%
%% ------------------------------------------------------------------------ %%
% The body of the article must start with a \begin{article} command
%
\clearpage
\vspace{10mm}
\textbf{Key Points:}\\

@@

The Antarctic ice shelves are characterized by large infrequent calving events where massive pieces of the ice shelves break off, to create giant tabular icebergs. Observational estimates suggest that over the past 30 years approximately half of Antarctic ice-shelf decay is due to iceberg calving, while the other half occurs through ice-shelf melting \citep{Depoorter2013, Rignot2013}. After calving, icebergs slowly drift away from their mother glaciers, often becoming stuck in sea ice, or become grounded along the Antarctic coast \citep{Lichey2001, Dowdeswell2007}. Giant tabular icebergs extend deep into the water column, and have to potential to disrupt ocean circulation patterns for months or even years after calving \citep{Robinson2012, Stern2015}. Iceberg sizes appear to fit a power law distribution, with the majority of icebergs being small (L$<$ 1 km), while the far less numerous giant tabular icebergs (L $>$ 5km) account for more than 90$\%$ of the icebergs mass \citep{Tournadre2016}.

Freshwater fluxes from iceberg melt impact ocean hydrography, influencing sea-ice production and bottom-water formation \citep{Arrigo2002, Robinson2012, Nicholls2009}. The long distances traveled by giant icebergs before melting means that their meltwater impact is often felt hundreds of kilometers away from their calving origins \citep{Stern2016}. Melt water injection (and the accompanying upwelling) from icebergs can also influence biological productivity by providing nutrients to the surface ocean or changing sea ice conditions \citep{Arrigo2002, Vernet2012, Biddle2015}. The increased productivity associated with free-floating tabular icebergs has been linked with local increases in ocean carbon uptake, potentially large enough to be a significant fraction of the Southern Ocean carbon sequestration \citep{Smith2007}.
%either increasing it by providing nutrients to the surface ocean, or decreasing by changing sea ice conditions

In recent years, there has been an increased interest in iceberg drift and decay. This surge of interest has been driven by (i) the need to understand polar freshwater cycles in order to create realistic climate forecasts and sea level projections \citep{Silva2006, Shepherd2007, Rignot2013}, (ii) the increased use of high-latitudes iceberg-filled waters for shipping lanes and offshore hydrocarbon exploration \citep{Pizzolato2014, Unger2014, Henderson2016}. The increased interest in icebergs has led to the development of numerical models of iceberg drift and decay \citep{Mountain1980, Bigg1997, Gladstone2001, Kubat2005}, some of which have been included in global General Circulation Models \citep{Martin2010, Marsh2015}. These iceberg models treat icebergs as Lagrangian point particles, which are advected by the flow, and melt according to certain parameterizations for icebergs melt. Since icebergs are treated as point particles, the iceberg models are mostly suitable for modeling smaller icebergs drifting across large distances, and as such iceberg models have mostly been used to represent icebergs smaller than 3.5km on a global scale \citep{Jongma2009, Martin2010, Marsh2015}.
These models are not suitable for modeling larger tabular icebergs (or smaller icebergs at fine scales), where the size and structure of the iceberg may be an important feature in determining their drift and decay \citep{Stern2016}. They also are not suitable for studying the local effects that icebergs have on the surrounding ocean. For this reason, tabular icebergs (L $>$ 5km) are currently not represented in the iceberg models used in climate models, even though observations suggest that tabular icebergs account for  the vast majority of the total Southern Hemisphere icebergs mass \citep{Tournadre2016}. Similarly, operational iceberg models are often inadequate for projecting iceberg drift at fine scales \citep{Turnbull2015}.

The goal of this study is to develop a new iceberg model where tabular icebergs are explicitly resolved in the ocean. In this model, icebergs are no longer treated as point particles that feel the ocean at one point, but rather the icebergs have structure, depress the ocean surface, and can feel the ocean across multiple ocean grid cells. To do this, we create tabular icebergs out of many Lagrangian elements which are `bonded' together by numerical bonds. The numerical bonds hold the ice elements together and allow a collection elements to move as a unit. By manually breaking these bonds, we can simulate iceberg calving, allowing us to study the ocean response to giant iceberg detaching from an ice shelf (e.g.: Figure \ref{fig:Spread_area} and \ref{fig:SST_Collapse}) or the fracturing of a large iceberg into smaller pieces.

The discrete element model (DEM) developed in this study is referred to at the Lagrangian Tabular Iceberg Model (LTAM). 
%The LTAM model is coupled to the MOM6 numerical ocean model \citep{Hallberg2013}. 
Section 2 gives a description of many of the key aspects of the LTAM. Since this model is a new approach to modeling iceberg, we present many of the technical elements involved in constructing the model. In Section 3 and 4, we demonstrate the capabilities of the model by simulating a tabular iceberg detaching from an idealized ice shelf.

@@

\section{Model description}
%, which can be used to model small icebergs, tabular icebergs, sea ice and ice shelf cavities
%In this section we describe many key elements of the LTAM.  
The LTAM is a Lagrangian particle-based model (or DEM model) in that the objects of the model are Lagrangian elements. Each element represents a mass of ice which is floating in the ocean, and has a position, velocity, mass, and a set of dimensions, which can evolve in time. The motion of each element is determined by a momentum equation which is solved in the (Lagrangian) reference frame of the element. The elements are forced by oceanic and atmospheric forces, which are provided by the user, or are determined by coupling the LTAM to an ocean/atmosphere model. The elements also interact with one another and can be bonded together to form larger structures. The angular momentum of the elements is not model explicitly, instead rotational motion of larger structures emerge as a consequence of bond orientation. In different contexts, the LTAM elements can be thought to represent individual icebergs, sea ice flows, or, when the elements are bonded together, they can represent larger structures such as tabular icebergs or ice shelves.
%In this latter case, the LTAM model can be categorized as a discrete element model (DEM). 
% in the sense that large structures are constructed from smaller independent Lagrangian elements.

@@

\subsection{Equations of motion}
The elements drift in the ocean, forced by atmosphere, ocean and sea-ice drag forces, as well as the Coriolis force and a force due to the sea surface slope.  When these ice elements move alone (without interactions with other elements), they can be thought of as representing individual (or clusters of) small icebergs, and follow the same equations described in the iceberg drift model of \citet{Martin2010} (based on the equations outlined in \citet{Bigg1997, Gladstone2001}).

@@

To aid in notation, we define a bond matrix $B_{ij}$ such that $B_{ij}=1$ if elements i and j are bonded together and  $B_{ij}=0$ otherwise. Using this notation, the interactive force $(\vec{F_{IA}})_{ij}$ on an element i by an element j is given by
\begin{equation}
\label{F_interactione}
(\vec{F_{IA}})_{ij}= \left\{%
  \begin{array}{ll}
   (\vec{F_{e}})_{ij} + (\vec{F_{d}})_{ij}
    & \mbox{if $d_{ij} \leq L_{ij} $ }\\
 %     & \mbox{if $|x_{i} -x_{j}| \leq \frac{1}{2}(D_{i}+D_{j}) $ }\\
    (\vec{F_{e}})_{ij} + (\vec{F_{d}})_{ij}
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=1 $ }\\
       & \mbox{if $d_{ij} >  L_{ij}  \textrm{ and } B_{ij}=1 $ }\\
      0
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=0 $ }.\\
      & \mbox{if $d_{ij} > L_{ij} \textrm{ and } B_{ij}=0 $ }.\\
  \end{array}
\right.
\end{equation}
$(\vec{F_{e}})_{ij}$ and $(\vec{F_{d}})_{ij}$ are the elastic and frictional damping components of the interactive force between elements i and j.
The elastic force $(\vec{F_{e}})_{ij}$ between elements is given by
\begin{equation}
\label{F_elastic}
(\vec{F_{e}})_{ij}= -\kappa_{e}  \bigg( d_{i,j}-L_{i,j} \bigg) T_{i,j} \vec{r},
\end{equation}
where $\vec{r}= \frac{(\vec{x_{i}} -\vec{x_{j}})}{|\vec{x_{i}} -\vec{x_{j}}|}$ is the directional unit vector between the position of element i and j, and $\kappa_{e}$ is the spring constant, and $T_{i,j}$ is the minimum of the thickness of elements i, j. We choose $(\vec{F_{b}})_{ij}$ proportional to $T_{i,j}$ (the minimum thickness of elements i and j) so the formulation of interactive forces has the desired property that it obeys Newton's 3rd Law (i.e.: $(\vec{F}_{IA})_{ij}=-(\vec{F}_{IA})_{ji}$).
%This is in contrast to some other DEM models in the literature (e.g. \citet{Li2014}). 
The minimum thickness is preferred to the average thickness, since this means that the for two bonded elements a fixed distance apart, acceleration due to elastic forces is bounded, even when the thickness of one of the elements approaches zero. 
%
%Since ice elements are able to melt entirely, the repulsive forces between elements need to be carefully defined so that they are stable even for very small elements with mass approaching zero. We define the elastic repulsive force $(F_{r})_{ij}$ between elements i and j as
%\begin{equation}
%\label{F_elastic}
%(F_{r})_{ij}= -\kappa_{r} A_{o}(i,j) T_{i,j}\vec{r},
%\end{equation}
%where $A_{o}(i,j)$ is the overlap area of the two circles centered at $\vec{x_{i}}$ and $\vec{x_{j}}$, with diameter $D_{i}$ and $D_{j}$ (see Figure \ref{fig:Interactive_forces}), respectively, and $\kappa_{r}$ is the repulsive spring constant. We choose $(F_{r})_{ij}$ to be proportional to the overlap area between the interactive circles surrounding each element so that the acceleration due to this repulsive force is bounded, even as the element mass approaches zero.   This consideration is not needed for the bonded force, since element bonds will be broken when elements are smaller than a threshold size. 
%[Should we use a bonded force given by  $(F_{b})_{ij}= -\kappa_{b}  \bigg( d_{i,j}-L_{i,j}) \bigg)^{2} T_{i,j} \vec{r}$ rather? Come back to this.]

@@

The effectiveness of the repulsive forces can be seen in Figure \ref{fig:Regular_Uncoupled_sim}, which shows an uncoupled (ice only) simulation where ice elements drift westward into a bay, and eventually come to rest with minimal overlap between elements. The effectiveness of the numerical bonds is demonstrated in Figure \ref{fig:Uncoupled_sim}, where tabular icebergs (constructed from many ice elements bonded together) and individual icebergs (unbonded elements) drift together towards a convex coast line. When the tabular icebergs arrive at the coast, they bump into the coastline and begin to rotate, influencing the paths of the other icebergs. In this example we see that modeling large structures using small elements bonded together, allows us to achieve large-scale structure and rotational motion, without having to include an equation for the angular momentum of the elements (as discussed in \citet{Jakobsen2001}). Animations of these simulations can be found in the supplementary materials.

@@

@@

\subsubsection{Time stepping}
The ice elements in the LTAM model are advected using a semi-implicit velocity Verlet time-stepping scheme. The velocity verlet time stepping scheme is commonly used in DEM models in video games because it is computational efficient and has desirable stability properties \citep{Jakobsen2001}. This time stepping scheme was preferred to the Runge-Kutta  4, which was used in the iceberg model of \citet{Martin2010} since, the Verlet time stepping only requires one calculation of the interactive forces per time step (while the Runge-Kutta scheme requires the interactive forces to be calculated four times). Since the calculation of the interactive forces is one of the most computationally expensive part of the algorithm, the Verlet scheme leads to a significant increase in the speed of the model. We note that the Verlet scheme used in the LTAM contains a small modification of the original (fully explicit) velocity Verlet time stepping scheme in that damping terms are treated implicitly (which increases the numerical stability). The details of this time stepping schemed are outline in Appendix B.

@@

 \section{Experiment Setup}
The introduction of Lagrangian elements, numerical bonds and interpolation schemes between the Eularian and Lagrangian grids (discussed in Section 2) means that we now have the tools to model giant tabular icebergs submerged in the ocean. We demonstrate this capability by performing a simulation of a tabular iceberg drifting away from an idealized ice shelf. This simulation is performed using the LTAM model coupled to the MOM6 ocean model \citep{Hallberg2013}. 
%This section describes the experimental setup of these numerical experiments. The results are shown in Section 4.

 \subsection{Model configuration}
We initialized our simulation using the experimental setup created for the Marine Ice Ocean Modeling Inter-comparion Project (MISOMIP) \citep{Asay-Davis2016}.
The configuration consists of an idealized ice shelf in a rectangular domain. The domain is $L_{x}=80$km wide and $L_{y}=480$km long, and contains an ice shelf which is grounded on the south side of the domain and has an ice front at y=650 km. The ice thickness and bottom topography of this setup are shown in Figure \ref{fig:ISOMIP_mass_and_topog}a and \ref{fig:ISOMIP_mass_and_topog}b respectively, with the grounding line position drawn in for reference. The configuration is the same as that of the Ocean0 setup in the MISOMIP, with three changes made:
\begin{enumerate}
\item The `calving criteria' used in the MISOMIP study (which states that all points in the ice shelf with thickness less than 100m are set to zero thickness) has not been used.
\item The ice shelf has been thickened on the flanks of the domain, so that the latitude of the grounding line increases away from the center of the ice shelf.
\item The ice shelf is configured to be symmetric about its meridional center line ($x=\frac{L_{x}}{2}$). This was achieved by using the average of the left and right flanks of the ice-shelf thickness. 
\end{enumerate}
These three changes were made in order to make the circulation beneath the ice shelf easier to interpret.

%\textbf{Initializing Lagrangian elements:}\\
\subsection{Initializing Lagrangian elements:}
The idealized ice shelf is constructed out of Lagrangian ice elements.
Ice elements are set to be hexagonal and are arranged in a regular staggered lattice (as discussed in Section 2.3). The sides of the gridded hexagons are initialized with length $S=0.98$ km. Recall that the element size must be smaller than the ocean grid spacing (i.e.: $2S<\Delta x$ = 2 km). 
Gaps along the boundaries are filled in using smaller elements so that the total ice-shelf area is preserved. The initial mass of the ice elements is determined by a preprocessing inversion step, which is the inverse of the 'mass-spreading' interpolation procedure discussed in Section 2.3.

Figure \ref{fig:ISOMIP_mass_and_topog}c shows what the ice-shelf draft would be if the draft were calculated from the mass of elements in each ocean grid cell without spreading an element's mass across neighboring cells (i.e.: treating elements as point masses). Figure \ref{fig:ISOMIP_mass_and_topog}b shows the draft after spreading the mass across grid cells. When the mass spreading interpolation scheme is not used, grid artifacts seen in the ice-shelf draft (Figure \ref{fig:ISOMIP_mass_and_topog}c). The grid artifacts are much reduced when the mass spreading interpolation is used (Figure \ref{fig:ISOMIP_mass_and_topog}b).

\subsection{Ocean model setup}
The LTAM model is coupled to the MOM6 ocean model.
The ocean model is run using vertical coordinate system which is a hybrid between a sigma-level and a z-level coordinate. In particular, model layers bend underneath the ice shelf as they would in a sigma-coordinate model, but collapse to zero thickness when they intersect with bottom topography, as they would in a z-level model. The coordinate system was achieve using ALE regridding-remapping scheme \citep{White2009}. The model uses a horizontal resolution of 2km, and 72 vertical layers.
All simulation were repeated using the ocean model run layered mode (results were similar and are not presented here).
%(select results are shown in the supplementary material). 
Ocean parameters are as specified in the MISOMIP configuration \citep{Asay-Davis2016}, and are shown in Table 1. The simulation is initially at rest, with horizontally uniform initial ocean temperature and salinity profiles that vary linearly between specified surface and bottom values. The initial surface temperature and salinity are  $T_{t}=-1.9^{\circ}$C and $S_{t}=33.8$ psu, respectively. The initial temperature and salinity at the depth $H_{ocean}$ are $T_{b}=1.0^{\circ}$C and $S_{b}=34.7$ psu. The maximum ocean depth is $H_{ocean}=720$ m.  A sponge layer is used on the northern boundary, which relaxes back to the initial temperature and salinity with a relaxation time scale of $T_{sponge}=0.1$ days. Melting is set to zero for ocean cells where the ocean column thickness is less than 10m to avoid using more energy to melt ice than is present in the water column.

>>>>

\begin{abstract}
The calving of giant icebergs from the Antarctic ice shelves accounts for approximately half of Antarctic ice-shelf decay. After calving these tabular icebergs drift across large distances, altering regional ocean circulation, bottom-water formation, sea-ice production and rates biological primary productivity in the icebergs' vicinity.
However, despite their importance, the current generation of numerical models is unable to represent giant tabular icebergs a physically realistic way.
In this study we develop novel framework to model giant tabular icebergs submerged in the ocean. In this framework, tabular icebergs are constructed out of Lagrangian elements that drift in the ocean, and are held together by numerical bonds.
By breaking these bonds, a tabular iceberg can be made to split into smaller pieces, or to break away from an ice shelf (also constructed from Lagrangian elements).
This paper provides a description of the Lagrangian tabular-iceberg model and discusses a number of the technical elements involved its development. 
We demonstrate the capabilities of the model by modeling a tabular iceberg drifting away an idealized ice shelf.
\end{abstract}
%% ------------------------------------------------------------------------ %%
%
%  BEGIN ARTICLE
%
%% ------------------------------------------------------------------------ %%
% The body of the article must start with a \begin{article} command
%
\clearpage
\vspace{10mm}
\textbf{Key Points:}\\

@@

The Antarctic ice shelves are characterized by large infrequent calving events where massive pieces of the ice shelves break off, to create giant tabular icebergs. Observational estimates suggest that over the past 30 years approximately half of Antarctic ice-shelf decay is due to iceberg calving, while the other half occurs through ice-shelf melting \citep{Depoorter2013, Rignot2013}. Iceberg sizes appear to fit a power law distribution, with the majority of icebergs being small (L$<$ 1 km), while the far less numerous giant tabular icebergs (L $>$ 5km) account for more than 90$\%$ of the icebergs mass \citep{Tournadre2016}.

After calving, icebergs slowly drift away from their mother glaciers, often becoming stuck in sea ice, or grounding along the Antarctic coast \citep{Lichey2001, Dowdeswell2007}. Giant tabular icebergs extend deep into the water column, and have to potential to disrupt ocean circulation patterns for months or even years after calving \citep{Robinson2012, Stern2015}.
The freshwater flux from iceberg melt impacts ocean hydrography around the iceberg, influencing sea-ice production and bottom-water formation \citep{Arrigo2002, Robinson2012, Nicholls2009}. The long distances traveled by giant icebergs before melting means that their meltwater impact is often felt hundreds of kilometers away from their calving origins \citep{Stern2016}. Meltwater injection (and the accompanying upwelling) from icebergs can also influence biological productivity by providing nutrients to the surface ocean or changing sea ice conditions \citep{Arrigo2002, Vernet2012, Biddle2015}. The increased productivity associated with free-floating tabular icebergs has been linked with local increases in ocean carbon uptake, potentially large enough to be a significant fraction of the Southern Ocean carbon sequestration \citep{Smith2007}.

%either increasing it by providing nutrients to the surface ocean, or decreasing by changing sea ice conditions

In recent years, there has been an increased interest in iceberg drift and decay. This surge of interest has been driven by (i) the need to understand polar freshwater cycles in order to create realistic climate forecasts and sea level projections \citep{Silva2006, Shepherd2007, Rignot2013}, (ii) the increased use of high-latitudes iceberg-filled waters for shipping lanes and offshore hydrocarbon exploration in the Arctic \citep{Pizzolato2014, Unger2014, Henderson2016}. The increased interest in icebergs has led to the development of numerical models of iceberg drift and decay \citep{Mountain1980, Bigg1997, Gladstone2001, Kubat2005}, some of which have been included in global General Circulation Models \citep{Martin2010, Marsh2015}. These iceberg drift models treat icebergs as Lagrangian point particles, which are advected by the flow, and melt according to certain parameterizations for icebergs melt. Since icebergs are treated as point particles, iceberg drift models are mostly suitable for modeling smaller icebergs drifting across large distances, and as such these models have mostly been used to represent icebergs smaller than 3.5km on a global scale \citep{Jongma2009, Martin2010, Marsh2015}.

Point-particle iceberg drift models are less suitable for modeling larger tabular icebergs, where the size and structure of the iceberg may be an important feature in determining their drift and decay \citep{Stern2016}. They also are not suitable for studying the local effects that icebergs have on the surrounding ocean, or the small scale processes that influence iceberg melt and decay \citep{Wagner2014, Stern2015}. For this reason, tabular icebergs (L $>$ 5km) are currently not represented in the iceberg drift models used in climate models, even though observations suggest that tabular icebergs account for  the vast majority of the total Southern Hemisphere iceberg mass \citep{Tournadre2016}. 
%Similarly, operational iceberg models are often inadequate for projecting iceberg drift at fine scales \citep{Turnbull2015}.

The goal of this study is to develop a new iceberg model where tabular icebergs are explicitly resolved in the ocean. In this model, icebergs are no longer treated as point particles that interact with the ocean at a single location, but rather icebergs are given structure, so that they interact with the ocean across multiple ocean grid cells and depress the ocean surface over a wide area. To do this, we create tabular icebergs out of many Lagrangian elements which are `bonded' together by numerical bonds. The numerical bonds hold the ice elements together and allow a collection elements to move as a unit. By manually breaking these bonds, we can simulate iceberg calving, allowing us to study the ocean response to giant iceberg detaching from an ice shelf (e.g.: Figure \ref{fig:Spread_area} and \ref{fig:SST_Collapse}) or the fracturing of a large iceberg into smaller pieces.

The discrete element model (DEM) developed in this study is referred to at the Lagrangian Tabular Iceberg Model (LTAM). 
%The LTAM model is coupled to the MOM6 numerical ocean model \citep{Hallberg2013}. 
Section 2 gives a description of many of the key aspects of the LTAM. Since this model is a new approach to modeling icebergs, we present many of the technical elements involved in constructing the model. In section 3 and 4, we demonstrate the capabilities of the model by simulating a tabular iceberg detaching from an idealized ice shelf.

@@

\section{Model description}
%, which can be used to model small icebergs, tabular icebergs, sea ice and ice shelf cavities
%In this section we describe many key elements of the LTAM.  
The LTAM is a Lagrangian particle-based model (or DEM model) in that the objects of the model are Lagrangian elements. Each element represents a mass of ice that is floating in the ocean, and has a position, velocity, mass, and a set of dimensions, which can evolve in time. The motion of each element is determined by a momentum equation which is solved in the (Lagrangian) reference frame of the element. The elements are forced by oceanic and atmospheric forces, which are provided by the user, or are determined by coupling the LTAM to an ocean/atmosphere model. The elements also interact with one another and can be bonded together to form larger structures. The angular momentum of the elements is not modeled explicitly; instead rotational motion of larger structures emerge as a consequence of bond orientation. In different contexts, the LTAM elements can be thought to represent individual icebergs, sea ice flows, or, when the elements are bonded together, they can represent larger structures such as tabular icebergs or ice shelves.
%In this latter case, the LTAM model can be categorized as a discrete element model (DEM). 
% in the sense that large structures are constructed from smaller independent Lagrangian elements.

@@

\subsection{Equations of motion}
The elements drift in the ocean, forced by atmosphere, ocean and sea-ice drag forces, as well as the Coriolis force and a force due to the sea surface slope.  When these ice elements move alone (without interactions with other elements), they can be thought of as representing individual (or clusters of) small icebergs, and follow the same equations described in the iceberg drift model of \citet{Martin2010} (based on the equations outlined in \citet{Bigg1997} and \citet{Gladstone2001}).

@@

To aid in notation, we define a bond matrix $B_{ij}$ such that $B_{ij}=1$ if elements i and j are bonded together and  $B_{ij}=0$ otherwise. Using this notation, the interactive force $(\vec{F_{IA}})_{ij}$ on an element i by an element j is given by
\begin{equation}
\label{F_interactione}
(\vec{F_{IA}})_{ij}= \left\{%
  \begin{array}{ll}
   (\vec{F_{e}})_{ij} + (\vec{F_{d}})_{ij}
    & \mbox{if $d_{ij} \leq L_{ij} $ }\\
 %     & \mbox{if $|x_{i} -x_{j}| \leq \frac{1}{2}(D_{i}+D_{j}) $ }\\
    (\vec{F_{e}})_{ij} + (\vec{F_{d}})_{ij}
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=1 $ }\\
       & \mbox{if $d_{ij} >  L_{ij}  \textrm{ and } B_{ij}=1 $ }\\
      0
 %     & \mbox{if $|x_{i} -x_{j}| > \frac{1}{2}(D_{i}+D_{j}) \textrm{ and } B_{ij}=0 $ }.\\
      & \mbox{if $d_{ij} > L_{ij} \textrm{ and } B_{ij}=0 $ }.\\
  \end{array}
\right.
\end{equation}
$(\vec{F_{e}})_{ij}$ and $(\vec{F_{d}})_{ij}$ are the elastic and frictional damping components of the interactive force between elements i and j.
The elastic force $(\vec{F_{e}})_{ij}$ between elements is given by
\begin{equation}
\label{F_elastic}
(\vec{F_{e}})_{ij}= -\kappa_{e}  \bigg( d_{i,j}-L_{i,j} \bigg) T_{i,j} \vec{r},
\end{equation}
where $\vec{r}= \frac{(\vec{x_{i}} -\vec{x_{j}})}{|\vec{x_{i}} -\vec{x_{j}}|}$ is the directional unit vector between the position of element i and j, $\kappa_{e}$ is the spring constant, and $T_{i,j}$ is the minimum of the thickness of elements i, j. We choose $(\vec{F_{b}})_{ij}$ proportional to $T_{i,j}$ (the minimum thickness of elements i and j) so the formulation of interactive forces has the desired property that it obeys Newton's 3rd Law (i.e.: $(\vec{F}_{IA})_{ij}=-(\vec{F}_{IA})_{ji}$).
%This is in contrast to some other DEM models in the literature (e.g. \citet{Li2014}). 
The minimum thickness is preferred to the average thickness, since this means that the for two bonded elements a fixed distance apart, acceleration due to elastic forces is bounded, even when the thickness of one of the elements approaches zero. 
%
%Since ice elements are able to melt entirely, the repulsive forces between elements need to be carefully defined so that they are stable even for very small elements with mass approaching zero. We define the elastic repulsive force $(F_{r})_{ij}$ between elements i and j as
%\begin{equation}
%\label{F_elastic}
%(F_{r})_{ij}= -\kappa_{r} A_{o}(i,j) T_{i,j}\vec{r},
%\end{equation}
%where $A_{o}(i,j)$ is the overlap area of the two circles centered at $\vec{x_{i}}$ and $\vec{x_{j}}$, with diameter $D_{i}$ and $D_{j}$ (see Figure \ref{fig:Interactive_forces}), respectively, and $\kappa_{r}$ is the repulsive spring constant. We choose $(F_{r})_{ij}$ to be proportional to the overlap area between the interactive circles surrounding each element so that the acceleration due to this repulsive force is bounded, even as the element mass approaches zero.   This consideration is not needed for the bonded force, since element bonds will be broken when elements are smaller than a threshold size. 
%[Should we use a bonded force given by  $(F_{b})_{ij}= -\kappa_{b}  \bigg( d_{i,j}-L_{i,j}) \bigg)^{2} T_{i,j} \vec{r}$ rather? Come back to this.]

@@

The effectiveness of the repulsive forces can be seen in Figure \ref{fig:Regular_Uncoupled_sim}, which shows an uncoupled (ice only) simulation where ice elements drift westward into a bay, and eventually come to rest with minimal overlap between elements. The effectiveness of the numerical bonds is demonstrated in Figure \ref{fig:Uncoupled_sim}, where tabular icebergs (constructed from many ice elements bonded together) and individual icebergs (unbonded elements) drift together towards a convex coast line. When the tabular icebergs arrive at the coast, they bump into the coastline and begin to rotate, influencing the paths of the other icebergs. In this example we see that modeling large structures using small elements bonded together, allows us to achieve large-scale structure and rotational motion, without having to include an equation for the angular momentum of the elements (as discussed in \citet{Jakobsen2001}). Animations of these uncoupled simulations can be found in the supplementary materials.

@@

Finally, we note that since the tabular iceberg are explicitly resolved in the ocean, it is sufficient for each element to interact with ocean mixed layer (i.e. there is no need to manually embed the iceberg into the ocean by integrating ocean fields over the iceberg thickness, as in \citet{Merino2016}).

@@

\subsubsection{Time stepping}
The ice elements in the LTAM model are advected using a semi-implicit velocity Verlet time-stepping scheme. The velocity verlet time stepping scheme is commonly used in DEM models in video games because it is computational efficient and has desirable stability properties \citep{Jakobsen2001}. This time stepping scheme was preferred to the Runge-Kutta  4, which was used in the iceberg model of \citet{Martin2010} since, the Verlet time stepping only requires one calculation of the interactive forces per time step (while the Runge-Kutta scheme requires the interactive forces to be calculated four times). Since the calculation of the interactive forces is one of the most computationally expensive part of the algorithm, the Verlet scheme leads to a significant increase in the speed of the model. We note that the Verlet scheme used in the LTAM contains a modification of the original (fully explicit) velocity Verlet time stepping scheme in that damping terms are treated implicitly (which increases the numerical stability). The details of this time stepping schemed are outline in Appendix B.

@@

 \section{Experiment Setup}
The introduction of Lagrangian elements, numerical bonds and interpolation schemes between the Eularian and Lagrangian grids (discussed in Section 2) means that we now have the tools to model giant tabular icebergs submerged in the ocean. We demonstrate this capability by simulating a tabular iceberg drifting away from an idealized ice shelf. 
%This simulation is performed using the LTAM model coupled to the MOM6 ocean model \citep{Hallberg2013}. 
%This section describes the experimental setup of these numerical experiments. The results are shown in Section 4.

 \subsection{Model configuration}
We initialized our simulation using the experimental setup created for the Marine Ice Ocean Modeling Inter-comparion Project (MISOMIP) \citep{Asay-Davis2016}.
The configuration consists of an idealized ice shelf in a rectangular domain. The domain is $L_{x}=80$km wide and $L_{y}=480$km long, and contains an ice shelf which is grounded on the south side of the domain and has an ice front at y=650 km. The ice thickness and bottom topography of this setup are shown in Figure \ref{fig:ISOMIP_mass_and_topog}a and \ref{fig:ISOMIP_mass_and_topog}b respectively, with the grounding line position drawn in for reference. The configuration is the same as that of the Ocean0 setup in the MISOMIP, with minor changes to the ice-shelf geometry (see Appendix D for details).
% (see supplementary materials for details).
%
%with three changes made:
%\begin{enumerate}
%\item The `calving criteria' used in the MISOMIP study (which states that all points in the ice shelf with thickness less than 100m are set to zero thickness) has not been used.
%\item The ice shelf has been thickened on the flanks of the domain, so that the latitude of the grounding line increases away from the center of the ice shelf.
%\item The ice shelf is configured to be symmetric about its meridional center line ($x=\frac{L_{x}}{2}$). This was achieved by using the average of the left and right flanks of the ice-shelf thickness. 
%\end{enumerate}
%These three changes were made in order to make the circulation beneath the ice shelf easier to interpret.

%\textbf{Initializing Lagrangian elements:}\\
\subsection{Initializing Lagrangian elements:}
The idealized ice shelf is constructed out of Lagrangian ice elements.
Ice elements are hexagonal and are arranged in a regular staggered lattice (as discussed in Section 2.3). The sides of the gridded hexagons are initialized with length $S=0.98$ km. %Recall that the element size must be smaller than the ocean grid spacing (i.e.: $2S<\Delta x$ = 2 km). 
Gaps along the boundaries are filled in using smaller elements so that the total ice-shelf area is preserved. The initial mass of the ice elements is determined by a preprocessing inversion step, which is the inverse of the `mass-spreading' interpolation procedure discussed in Section 2.3. 
The pre-interpolation ice draft (treating elements as point masses) contains large grid artifacts (Figures \ref{fig:ISOMIP_mass_and_topog}c). These grid artifacts are much reduced after the mass-spreading interpolation is applied (Figure \ref{fig:ISOMIP_mass_and_topog}b).

%Figure \ref{fig:ISOMIP_mass_and_topog}c shows what the ice-shelf draft would be if the draft were calculated from the mass of elements in each ocean grid cell without spreading an element's mass across neighboring cells (i.e.: treating elements as point masses). Figure \ref{fig:ISOMIP_mass_and_topog}b shows the draft after spreading the mass across grid cells. When the mass spreading interpolation scheme is not used, grid artifacts seen in the ice-shelf draft (Figure \ref{fig:ISOMIP_mass_and_topog}c). These grid artifacts are much reduced when the mass spreading interpolation is used (Figure \ref{fig:ISOMIP_mass_and_topog}b).

\subsection{Ocean model setup}
The LTAM model is coupled to the MOM6 ocean model.
The ocean model is run using a vertical coordinate system which is a hybrid between a sigma-level and a z-level coordinate. In particular, model layers bend underneath the ice shelf as they would in a sigma-coordinate model, but collapse to zero thickness when they intersect with bottom topography, as they would in a z-level model. The coordinate system was achieve using ALE regridding-remapping scheme \citep{White2009}. The model uses a horizontal resolution of 2km, and 72 vertical layers.
All simulation were repeated using the ocean model run layered mode (results were similar and are not presented here).
%(select results are shown in the supplementary material).

Ocean parameters are as specified in the MISOMIP configuration \citep{Asay-Davis2016}, and are shown in Table 1. The simulation is initially at rest, with horizontally uniform initial ocean temperature and salinity profiles which vary linearly between specified surface and bottom values: $T_{top}=-1.9^{\circ}$C, $T_{bottom}=1.0^{\circ}$C, $S_{top}=33.8$ psu, $S_{bottom}=34.7$. The maximum ocean depth is $H_{ocean}=720$ m.  A sponge layer is used on the northern boundary, which relaxes back to the initial temperature and salinity with a relaxation time scale of $T_{sponge}=0.1$ days. Melting is set to zero for ocean cells where the ocean column thickness is less than 10m to avoid using more energy to melt ice than is present in the water column.
